---
title: "Possible Analyses for Revisions"
format: 
  html:
    code-fold: true
bibliography: revisions_analysis.bib
---

```{r}
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(mgcv)
library(gratia)
library(DHARMa)
library(kableExtra)

theme_set(theme_minimal(base_size = 12))

data_folder <- "../data"
data_files <- paste0(data_folder,c("/data_1.csv","/data_2_temporal.csv","/data_to_map.csv","/data_to_map_21.csv"))

data_1 <- read_csv(data_files[1])
data_2_temporal <- read_csv(data_files[2])
data_to_map <- read_csv(data_files[3])
data_to_map_21 <- read_csv(data_files[4])

data_1 <- data_1 %>%
  mutate(pop_log = log10(pop),
         infested = i21)

data_2_temporal <- data_2_temporal %>%
  mutate(pop_log = log10(pop),
         year_zero = year - min(year))

data_to_map_166 <- data_to_map %>%
  filter(fips %in% data_2_temporal$fips)

data_14_21 <- data_2_temporal %>%
  select(-latitude,-longitude,-region,-sid,-e_dist)
```


## Overview

Using data collected on infestation by the spotted laternfly across 166 counties in the Mid-Atlantic region of the United States, we are interested to know about factors that likely contribute to spread of the organism. Here, we focus on factors related to human activity and that can be directly targeted for control or mitigation measures. The factors we focus on are:

* The number of businesses designated as garden centers in a county.

* The number of two-digit interstate highways that transect a county.

Further, we also examine the influence of the county population as previous work has established this as an important contributor to lanternfly spread. 

To develop an appropriate model for examining the impact of the above factors, it is necessary to control for spatial autocorrelation. If the infestation is modeled over multiple years, both spatial and temporal autocorrelation need to be addressed. Two common approaches to dealing with such autocorrelation in ecological modeling and statistics are generalized additive models [@wood2003; @wood2017generalized] and blocked-bootstrapping [@efron2022computer]. 

## The Data

The first few rows of the data collected appears as

```{r}
data_14_21 %>%
  select(infested,year,county,state,nis,gc,pop) %>%
  head() %>%
  kable()
```

For analysis, we add to this data the boundaries for all included counties.   

```{r}
data_to_map_166 <- data_to_map %>%
  filter(fips %in% data_14_21$fips) %>%
  select(fips,long,lat,group)

dftm <- data_14_21 %>%
  right_join(.,data_to_map_166,by="fips",
             relationship = "many-to-many")

```

@fig-figure-1 displays the 2021 infestation status for the 166 Mid-Atlantic region counties in 2021:

```{r}
#| label: fig-figure-1
#| fig-cap: Spotted laternfly infested counties for 2021. 

dftm %>%
  filter(year == "2021") %>%
  ggplot() +
  geom_polygon(mapping = aes(long, lat, group=group,fill=factor(infested)),color="black",
               linewidth=0.2)  +
  scale_fill_discrete(name="Infested",type=c("white","lightgreen")) + 
  labs(x="",y="",title = "2021 Infestations")
```


## Initial Exploratory Plots

```{r}
#| label: fig-figure-2
#| fig-cap: Here's the caption

dftm %>%
  filter(year == "2021") %>%
  ggplot() +
  geom_polygon(mapping = aes(long, lat, group=group,fill=factor(infested)),color="black",
               linewidth=0.2)  +
  scale_fill_discrete(name="Infested",type=c("white","lightgreen")) + 
  geom_point(aes(longitude_num,latitude_num,color=nps),size=2) + 
  scale_color_continuous(low = "white", 
                        high = "darkblue",
                        limits = c(0,5)) + 
  labs(x="",y="",title = "2021 Infestations per NPS")
```


```{r}
#| label: fig-figure-3
#| fig-cap: Here's the caption

dftm %>%
  filter(year == "2021") %>%
  ggplot() +
  geom_polygon(mapping = aes(long, lat, group=group,fill=factor(infested)),color="black",
               linewidth=0.2)  +
  scale_fill_discrete(name="Infested",type=c("white","lightgreen")) + 
  geom_point(aes(longitude_num,latitude_num,color=nis),size=2) + 
  scale_color_continuous(low = "white", 
                         high = "darkblue",
                         limits = c(0,3)) + 
  labs(x="",y="",title = "2021 Infestations per NIS")
```


```{r}
#| label: fig-figure-4
#| fig-cap: Here's the caption

dftm %>%
  filter(year == "2021") %>%
  ggplot() +
  geom_polygon(mapping = aes(long, lat, group=group,fill=factor(infested)),color="black",
               linewidth=0.2)  +
  scale_fill_discrete(name="Infested",type=c("white","lightgreen")) + 
  geom_point(aes(longitude_num,latitude_num,color=gc),size=2) + 
  scale_color_continuous(low = "white", 
                         high = "darkblue",
                         limits = c(0,41)) + 
  labs(x="",y="",title = "2021 Infestations per GC")
```



```{r}
#| label: fig-figure-5
#| fig-cap: Here's the caption

dftm %>%
  filter(year == "2021") %>%
  ggplot() +
  geom_polygon(mapping = aes(long, lat, group=group,fill=factor(infested)),color="black",
               linewidth=0.2)  +
  scale_fill_discrete(name="Infested",type=c("white","lightgreen")) + 
  geom_point(aes(longitude_num,latitude_num,color=pop_log),size=2) + 
  scale_color_continuous(low = "white", 
                         high = "darkblue",
                         limits = c(3,7)) + 
  labs(x="",y="",title = "2021 Infestations per Pop (log)")
```



## References

::: {#refs}
:::


:::{.callout-tip collapse="true"}
## Expand for Session Info
```{r}
#| echo: false


library(sessioninfo)
# save the session info as an object
pkg_sesh <- session_info(pkgs = "attached")

# get the quarto version
quarto_version <- system("quarto --version", intern = TRUE)

# inject the quarto info
pkg_sesh$platform$quarto <- paste(
  system("quarto --version", intern = TRUE), 
  "@", 
  quarto::quarto_path()
  )

# print it out
pkg_sesh
```

:::
