---
title: "Spread Paper Revisions"
subtitle: "Statistical Analysis"
author: "JMG"
format: 
  pdf:
    keep-tex: true
execute: 
  echo: false
  message: false
bibliography: revisions_analysis.bib
link-citations: true
link-bibliography: true
---

```{r}
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(mgcv)
library(gratia)
library(broom.mixed)
library(kableExtra)
library(DHARMa)
library(sjPlot)
library(sjlabelled)
library(sjmisc)
library(scales)

theme_set(theme_minimal(base_size = 10))

data_folder <- "../data"
data_files <- paste0(data_folder,c("/data_1.csv","/data_2_temporal.csv","/data_to_map.csv","/data_to_map_21.csv"))

data_1 <- read_csv(data_files[1])
data_2_temporal <- read_csv(data_files[2])
data_to_map <- read_csv(data_files[3])
data_to_map_21 <- read_csv(data_files[4])

data_1 <- data_1 %>%
  mutate(pop_log = log10(pop),
         infested = i21,
         is_presence = ifelse(is_presence == 0, "No","Yes"))

data_2_temporal <- data_2_temporal %>%
  mutate(pop_log = log10(pop),
         year_zero = year - min(year),
         is_presence = ifelse(is_presence == 0, "No","Yes"))

data_to_map_166 <- data_to_map %>%
  filter(fips %in% data_2_temporal$fips)

data_14_21 <- data_2_temporal %>%
  select(-latitude,-longitude,-region,-sid,-e_dist)

data_to_map_166 <- data_to_map %>%
  filter(fips %in% data_14_21$fips) %>%
  select(fips,long,lat,group)

dftm <- data_14_21 %>%
  right_join(.,data_to_map_166,by="fips",
             relationship = "many-to-many")
```

## Statistical analysis to determine relevance of factors

For a statistical test of our hypothesis, we have use R version 4.3.1 and package `mgcv` version 1.9 to fit a generalized additive model [@rcore; @wood2003; @wood2017generalized]. Specifically, we model the log odds of county infestation in 2021 with tensor product smoothing for longitude and latitude to control for spatial autocorrelation and parametric terms for presence/absence of two-digit interstate highway, number of garden centers, 2019 county population, and an interaction term for presence/absence of two-digit interstate highway and number of garden centers. To deal with issues of convergence and variables of different scale, we log transformed the population and normalized the number of garden centers. We used the packages `DHARMa` version 0.4.6 and `gratia` gratia version 0.8.1.34 for diagnostics to assess model assumptions  [@dharma2022 ; @gratia2023].

@fig-estimates shows the estimates on the odds-ratio scale for the parametric terms in our generalized additive logistic regression model while @tbl-mod-summ displays the values for the same estimated coefficients on the log-odds scale. The model results indicate that the data provides evidence for the increase in likelihood of infestation in a county in 2021 with the presence of interstate highways and larger populations. Additionally, our statistical analysis also suggests that the data provides evidence for the increase in likelihood of infestation in a county in 2021 with an increase in the number of garden centers only when there is also the presence of a two-digit interstate highway for that county. 

```{r}
#| code-fold: false

mod_form_21 <- infested ~ is_presence + std_gc + is_presence:std_gc + pop_log + 
   te(longitude,latitude)

gam_mod_21 <- gam(mod_form_21,data=data_1,
                      family = binomial(),
                      method = "REML")
```




```{r}
#| label: fig-estimates
#| fig-cap: Estimated coefficients for the linear predictors in the generalized additive model. Results are on the scale of odds-ratio. The likelihood of infestation for a county in 2021 is predicted to increase with the presence of interstate highways and larger populations. Interestingly, the likelihood of infestation for a county in 2021 is predicted to increase with an increase in the number of garden centers only when there is also the presence of a two-digit interstate highway for that county. 


gam_mod_21_ests <- plot_model(gam_mod_21)$data %>%
  tibble() %>%
  mutate(coeff = factor(c("IS Presence (Yes)","Garden Centers","Population","IS Presence (Yes) : Garden Centers"),levels=c("IS Presence (Yes) : Garden Centers","Population","Garden Centers","IS Presence (Yes)")))

gam_mod_21_ests %>%
  ggplot(aes(x=log(estimate),y=coeff)) + 
  geom_point(size=3) + 
  geom_errorbar(aes(xmin=log(conf.low), xmax=log(conf.high)), 
                width=0.0,
                 position=position_dodge(0.05),linewidth=1) +
  annotate(geom = "text", x = log(3.34)+0.1,y=4.2,label="p-value=0.054",size=2.5) +
  annotate(geom = "text", x = log(0.276),y=3.2,label="p-value=0.041",size=2.5) +
  annotate(geom = "text", x = log(5.37),y=2.2,label="p-value=0.046",size=2.5) +
  annotate(geom = "text", x = log(7.10),y=1.2,label="p-value=0.005",size=2.5) +
    scale_x_continuous(breaks = log(c(0.01,0.1,1.0,10.0,100.0)), 
                       labels = c(0.01,0.1,1.0,10.0,100.0),
                       limits = log(c(0.01,100.0))) + 
  labs(x = "Odds Ratio", y = "", title = "Parametric estimates for infestation") + 
  geom_vline(xintercept = 0.0,linetype="dashed")

```



```{r}
#| echo: false
#| label: tbl-mod-summ
#| tbl-cap: Summary table for estimates for linear terms in generalized additive model. Results are on the log-odds scale.

gam_mod_21_summ <- summary(gam_mod_21)

results_df <- tibble(Coefficient = c("Intercept","IS Presence - Yes","Garden Centers","Population (log)","IS Presence - Yes : Garden Centers"),
                     Estimate = gam_mod_21_summ$p.coeff,
                     SE = gam_mod_21_summ$se[1:5],
                     `p-value` = gam_mod_21_summ$p.pv)

results_df %>%
  kable()

```



## Supplementary analysis

@fig-figure-1 displays the 2021 infestation status for the 166 Mid-Atlantic region counties in 2021:

```{r}
#| echo: false
#| label: fig-figure-1
#| fig-cap: Counties in the defined Mid-Atlantic regions that were designated as infested by the spotted laternfly in 2021. 

dftm %>%
  filter(year == "2021") %>%
  ggplot() +
  geom_polygon(mapping = aes(long, lat, group=group,fill=factor(infested)),color="black",
               linewidth=0.2)  +
  scale_fill_discrete(name="Infested",type=c("white","lightgreen"),labels=c("No","Yes")) + 
  labs(x="Longitude",y="Latitude",title = "2021 Infestations")
```



@tbl-infestations displays the number of counties per year in the defined Mid-Atlantic region of 166 that were listed as infected.


```{r}
#| echo: false
#| label: tbl-infestations
#| tbl-cap: The number of counties per year in the defined Mid-Atlantic region of 166 that were listed as infected.

data_14_21 %>%
  group_by(year) %>%
  summarise(number_infested = sum(infested)) %>%
  rename(Year = year, `Number Infested` = number_infested) %>%
  kable()
```



```{r}
#| echo: false
#| include: false
#| label: fig-figure-2
#| fig-cap: Spotted laternfly infested counties for 2021 together with the numper of national parks in the county.

dftm %>%
  filter(year == "2021") %>%
  ggplot() +
  geom_polygon(mapping = aes(long, lat, group=group,fill=factor(infested)),color="black",
               linewidth=0.2)  +
  scale_fill_discrete(name="Infested",
                      type=c("white","lightgreen"),
                      labels=c("No","Yes")) + 
  geom_point(aes(longitude_num,latitude_num,color=nps),size=2) + 
  scale_color_continuous(low = "white", 
                        high = "darkblue",
                        limits = c(0,5)) + 
  labs(x="",y="",title = "2021 Infestations per NPS")
```


@fig-figure-3 shows the 166 counties in the defined Mid-Atlantic region together with their 2021 infestation status and the number of two-digit interstate highways that transect the county.

```{r}
#| echo: false
#| label: fig-figure-3
#| fig-cap: Spotted laternfly infested counties for 2021 together with the number of two-digit interstate highways transecting the counties.

dftm %>%
  filter(year == "2021") %>%
  ggplot() +
  geom_polygon(mapping = aes(long, lat, group=group,fill=factor(infested)),color="black",
               linewidth=0.2)  +
  scale_fill_discrete(name="Infested",
                      type=c("white","lightgreen"),
                      labels=c("No","Yes")) + 
  geom_point(aes(longitude_num,latitude_num,color=nis),size=2) + 
  scale_color_continuous(low = "white", 
                         high = "darkblue",
                         limits = c(0,3)) + 
  labs(x="Longitude",y="Latitude",title = "2021 Infestations per number of interstates",color="Interstates")
```


@fig-figure-4 shows the 166 counties in the defined Mid-Atlantic region together with their 2021 infestation status and the number of garden centers in the county.

```{r}
#| echo: false
#| label: fig-figure-4
#| fig-cap: Spotted laternfly infested counties for 2021 together with the number of garden centers in the counties.

dftm %>%
  filter(year == "2021") %>%
  ggplot() +
  geom_polygon(mapping = aes(long, lat, group=group,fill=factor(infested)),color="black",
               linewidth=0.2)  +
  scale_fill_discrete(name="Infested",type=c("white","lightgreen"),labels=c("No","Yes")) + 
  geom_point(aes(longitude_num,latitude_num,color=gc),size=2) + 
  scale_color_continuous(low = "white", 
                         high = "darkblue",
                         limits = c(0,41)) + 
  labs(x="Longitude",y="Latitude",title = "2021 Infestations per number of garden centers",
       color="Garden \n Centers")
```

@fig-figure-5 shows the 166 counties in the defined Mid-Atlantic region together with their 2021 infestation status and the county population on the log scale as estimated in 2019.

```{r}
#| echo: false
#| label: fig-figure-5
#| fig-cap: Spotted laternfly infested counties for 2021 together with the 2019 estimated population (log scale) for the counties.

dftm %>%
  filter(year == "2021") %>%
  ggplot() +
  geom_polygon(mapping = aes(long, lat, group=group,fill=factor(infested)),color="black",
               linewidth=0.2)  +
  scale_fill_discrete(name="Infested",type=c("white","lightgreen"),labels=c("No","Yes")) + 
  geom_point(aes(longitude_num,latitude_num,color=pop_log),size=2) + 
  scale_color_continuous(low = "white", 
                         high = "darkblue",
                         limits = c(3,7)) + 
  labs(x="",y="",title = "2021 Infestations per population (log)")
```


@fig-figure-6 shows the 166 counties in the defined Mid-Atlantic region together with the presence/absence of two-digit interstate highways and the number of garden center per county.


```{r}
#| echo: false
#| label: fig-figure-6
#| fig-cap: Spotted laternfly infested counties for 2021 together with the presence/absence of two-digit interstate highways and the number of garden center per county.

dftm %>%
  filter(year == "2021") %>%
  ggplot() +
  geom_polygon(mapping = aes(long, lat, group=group,fill=is_presence),color="black",
               linewidth=0.2)  +
  scale_fill_discrete(name="IS Presence",type=c("white","yellow")) + 
  geom_point(aes(longitude_num,latitude_num,color=gc),size=2) + 
  scale_color_continuous(low = "white", 
                         high = "darkblue",
                         limits = c(3,7)) + 
  labs(x="",y="",title = "Interstate presence and number of garden centers",
       color="Garden \n Centers")

```


### Infestation Trend Over Time


```{r}
#| code-fold: false

mod_form <- infested ~ is_presence + std_gc + is_presence:std_gc + pop_log + 
   s(year_zero,k=8) + te(longitude_num,latitude_num)

gam_mod <- gam(mod_form,data=data_14_21,
                      family = binomial(),
                      method = "REML")
```


````{r}
mod_smooth_estimates <- smooth_estimates(gam_mod) %>%
  filter(smooth == "s(year_zero)")

mod_smooth_estimates_der <- derivatives(gam_mod,term = "s(year_zero)")
```

[@gratia2023]

```{r}
#| label: fig-time
#| fig-cap: Partial effect plot for the log-odds of infestation over time from 2014 - 2021.


mod_smooth_estimates %>%
  ggplot(aes(x=year_zero + 2014,y=est)) + 
  geom_ribbon(aes(ymin = est - se,ymax = est + se),fill="lightgray") +
  geom_line(linewidth=1) +
  labs(x = "Year", y = "Partial effect")
```


```{r}
#| label: fig-time-deriv
#| fig-cap: Time rate of change (i.e., first derivative) of partial effect of the log-odds of infestation over time from 2014 - 2021. 

mod_smooth_estimates_der %>%
  ggplot(aes(x=data + 2014,y=derivative)) + 
  geom_ribbon(aes(ymin = derivative - se,ymax = derivative + se),fill="lightgray") +
  geom_line(linewidth=1) +
  labs(x = "Year", y = "Rate of change") + 
  ylim(c(0,3))
```


## References

::: {#refs}
:::
